---
title: "Homework assignment (Journal entry) - Docker"
output: html_notebook
---

```{r}
suppressMessages(library(DESeq2))
suppressMessages(library(pheatmap))
# create a 5 by 10 matrix of random integers
set.seed(0)
( m <- matrix(sample(1:1000,50), byrow=FALSE, ncol = 10, nrow = 5) )
# define column names as cond1, cond2, cond3, cond4, cond5, ctrl1, ctrl2, ctrl3, ctrl4, ctrl5
# define row names as gene1, gene2, gene3 ...
colnames(m) <- c("cond1","cond2","cond3","cond4","cond5",
                "ctrl1","ctrl2","ctrl3","ctrl4","ctrl5")
rownames(m) <- c("gene1","gene2","gene3","gene4","gene5")
View(m)
```


```{r}
# Compute the fold change for each gene.
foldchange <- function(row) {
  out <- 0
  for (i in 1:5) {
    out = out + ( row[i]/row[5+i] ) # hard-coded; compares 1 to 6, 2 to 7 etc.
  }
  return(out/5)
}
( m_fc <- apply(m, 1, foldchange) )
( m <- cbind(m,m_fc) )
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
